### 2018-02-21 reports/initial-proposal/initial-proposal.tex
Initial commit

%        File: initial-proposal.tex
%     Created: Tue Feb 20 10:00 PM 2018 E
% Last Change: Tue Feb 20 10:00 PM 2018 E
%
\documentclass[a4paper]{amsart}
\begin{document}

\title{APMTH 221 Final Project Proposal}
\author{Jiafeng (Kevin) Chen \and Francisco Rivera}
\date{\today}

\maketitle

\section{Problem Statement}

\end{document}

### 2018-02-21 reports/initial-proposal/initial-proposal.tex
Define problem statement in initial proposal

@@
 
+\section{Collaboration}
+
+Francisco and Kevin will work together to research the relevant optimization
+techniques helpful to calculating the Wasserstein distance quickly. Then, they
+will split up techniques, and through implementations of them, empirically (and
+when possible, theoretically) assess their performance. Time permitting, they
+will then attempt to create a new algorithm that performs competitively with
+state of the art benchmarks.
+
 \section{Problem Statement}
 
+If we are given two probability measures $\mu$ and $\nu$ on metric space
+$(M,d)$ with finite moments,
+\[ \int_M d(x,x_0)^p d\mu(x) < \infty \text{ and }\int_M d(x,x_0)^p d\nu(x) <
+\infty\]
+then we define the $p^\text{th}$ Wasserstein distance as
+\[ W_p(\mu, \nu) := \left( \inf_{\gamma \in \Gamma(\mu, \nu)} \int_{M \times
+M} d(x,y)^p \text{d} \gamma(x,y) \right)^{1/p}\]
+where $\Gamma(\mu, \nu)$ is the collection of all measures on $M \times M$ with
+marginals $\mu$ and $\nu$.
+
+This can also be thought of as the optimal transport problem. If we think of a
+probability distribution as a collection of ``mass'' at different points in $M$
+(with density of the mass proportional to the probability measure), then the
+Wasserstein distance captures the cost of transforming one distribution into the
+other in units of mass times distance.  This analogy is particularly
+interpretable if we have a discrete metric space, where probability measures can
+be viewed as point masses.
+
+Solving the optimization problem underpinning this metric is in general hard and
+remains an unsolved problem. We will aim to test the effectiveness of different
+optimization techniques to this problem, with particular focus on discrete
+metric spaces.
+
+\section{Motivation}
+
+Numerous important applications exist for the Wasserstein metric. 
+% TODO: what are these important applications?
+
+\section{Data}
+
+Because wish to test the effectiveness of Wasserstein distance calculations for
+arbitrary distributions, we will generate test distributions using simulation
+and do not need to concern ourselves with finding real data. When necessary, we
+will use standard distributions found in the literature for consistency.
+
+\section{Deliverables}
+
+We will implement optimization techniques to calculate Wasserstein distances and
+attempt a novel algorithm that performs competitively.
+
+\section{Next Steps}
+
+% TODO: what are we doing next?
+
 \end{document}

### 2018-02-21 reports/initial-proposal/initial-proposal.tex
Start working on future work
Suggest neural networks and optimization techniques

@@
 
+We have the following tentative ideas which we may or may not individually
+implement, but from which we will source our  work,
+
+\begin{itemize}
+
+\item We will use linear programs to calculate Wasserstein distances for
+discrete metric spaces. We will explore average complexity for randomly
+generated test cases.
+
+\item Training a neural network to hot-start an optimization algorithm for the
+discrete case.
+
+\item Exploring other relevant optimization techniques such as simulated
+annealing and genetic algorithms.
+
+\end{itemize}
+
 % TODO: what are we doing next?

### 2018-02-22 reports/initial-proposal/initial-proposal.tex
edits

--- 2018-02-21 reports/initial-proposal/initial-proposal.tex
@@
 %
-\documentclass[a4paper]{amsart}
+\documentclass{amsart}
+\usepackage{../am221}
+
+\usepackage[margin=1in]{geometry}
+\linespread{1.1}
+
 \begin{document}
@@
 
+\section{Collaboration}
+
+Francisco and Kevin will work together to research the relevant optimization
+techniques for efficiently computing Wasserstein distances and solving optimal transport problems. Then, they
+will split up techniques, and through implementations of them, empirically (and
+when possible, theoretically) assess their performance. Time permitting, they
+will then attempt to create a new algorithm that performs competitively with
+state of the art benchmarks.
+
 \section{Problem Statement}
 
+Given two probability distributions $p_X$ and $p_Y$, the \emph{$p$\textsuperscript{th} Wasserstein distance} is the following optimization problem: \[
+W_p(p_X,p_Y) = \pr{\inf_{(X,Y)} \br{ \mathbb{E}_{(X,Y)} \bk{d(X,Y)^p}: X\sim p_X, Y\sim p_Y }}^{1/p},
+\]
+where the minimum is taken over all joint distributions with marginals $p_X,p_Y$.
+The Wasserstein distance is a probabilistic view of a class of problems called \emph{optimal transport problems} \cite{peyre2017computational}.
+If we think of a
+probability distribution as a collection of ``mass'' at different points in $M$
+(with density of the mass proportional to the probability measure), then the
+Wasserstein distance captures the cost of transforming one distribution into the
+other in units of mass times distance.  This analogy is particularly
+interpretable if we consider discrete probability measures, which are collections of point masses.
+
+Solving the optimization problem underpinning this metric is in general hard and
+remains an unsolved problem. We will aim to test the effectiveness of different
+optimization techniques to this problem, with particular focus on discrete
+distributions.
+
+\section{Motivation}
+
+Numerous important applications exist for the Wasserstein metric.
+For instance, if we consider images as a discrete probabilistic distribution over $\mathbb{R}^2$,\footnote{For example, we may consider black-and-white images, where the probability mass is proportional to the pixel intensity.} the Wasserstein distance turns out to mimick human perceptions of similarity: Images that are similar to humans are close in Wasserstein distances.
+Other distances over distributions, such as the well-known Kullback-Leibler divergence, do not have such properties.
+Thus, Wasserstein distances are extremely important in computer vision, machine learning, and statistics.
+% TODO: what are these important applications?
+
+\section{Data}
+
+As this is a largely theoretical and methodological project, we shall rely on numerical simulation and analysis. We will generate test distributions using simulation
+and do not need to concern ourselves with finding real data. When necessary, we
+will use standard distributions found in the literature for consistency.
+
+\section{Deliverables}
+
+We will implement optimization techniques to calculate Wasserstein distances and
+attempt a novel algorithm that performs competitively. The deliverable will be a survey of recent literature and a report of our contributions.
+
+\section{Next Steps}
+
+First, we will review the current literature and familiarize ourselves with cutting-edge techniques in solving optimal transport problems.
+
+We have the following tentative ideas which we may or may not individually
+implement, but from which we will source our  work,
+
+\begin{itemize}
+
+\item We will use linear programs to calculate Wasserstein distances for
+discrete metric spaces.\footnote{This is the Kantorovich problem, see \cite{peyre2017computational}} We will explore average complexity for randomly
+generated test cases.
+
+\item Training a neural network to hot-start an optimization algorithm for the
+discrete case, following recent literature on the creative use of deep learning techniques for difficult optimization problems as in \cite{dutting2017optimal}.
+
+\item Exploring other relevant optimization techniques such as simulated
+annealing and genetic algorithms.
+
+\end{itemize}
+
+% TODO: what are we doing next?
+\bibliographystyle{plain}
+
+\bibliography{proposal.bib}
+
 \end{document}

### 2018-03-24 reports/progress-report/progress.tex
progress report theory 1/

\documentclass[11pt]{amsart}
\usepackage{../am221}
\linespread{1.2}
\title{Optimal Transport Problems}
\renewcommand{\b}{\mathbf}
\newcommand{\R}{\mathbb{R}}
\usepackage{bbm}
\newcommand{\one}{\mathbbm{1}}
\newcommand{\E}{\mathbb{E}}
\author{Jiafeng Chen\and Francisco Rivera}
\date{\today}

\theoremstyle{remark}
\newtheorem{rmk}{Remark}
\begin{document}
    \maketitle
    \section{Theoretical foundations} 
    Let $\b a \in \R^m_{\ge 0}$ and $\b b \in \R^n_{\ge 0}$ be two vectors of
    nonnegative real numbers such that $\b a^T \one = \b b^T \one = 1$, where
    $a_i$ and $b_j$ represents the amount of ``mass'' at different positions. 
    We are
    interested in a matrix $\b P \in \R^{m \times n}_{\ge 0}$ where $P_{ij}$
    prescribes
    the amount of mass flowing from $a_i$ to $b_j$. Naturally, we must have
    $\sum_j P_{ij} = a_i$ and $\sum_i P_{ij} = b_j$, so that mass is preserved.
    Let $\b C
    \in \R^{m\times n}$ be a \emph{cost matrix}, where $C_{ij}$ represents the
    cost of transporting one unit of $a_i$ to $b_j$. The \emph{Kantorovich
    optimal transport problem} is the following linear program: \begin{align*}
    \min_{\b P \in  \R^{m \times n}_{\ge 0}} \, & \sum_{i,j} P_{ij} C_{ij} \\
    \text{subjected to } & \b P \one = \b a \\
    & \b P^T \one = \b b
    \end{align*}

    \begin{rmk}[A probabilistic view]
    Let $X,Y$ be discrete random variables with support $\b
    x \in \mathcal X^m$ and $\b y \in \mathcal Y^n$, such that their marginal
    distributions are
    prescribed by $P(X = x_i) = a_i$ and $P(Y =
    y_j) = b_j$. Let $P_{ij} = P(X = x_i, Y = y_j)$, making $\b P = (P_{ij})_
    {ij}$ a joint
    distribution. Let $C: \mathcal X \times \mathcal Y \to \R$ be a cost
    function. Then the
    Kantorovich problem is equivalent to the following:\[
    \min_{\b P}\, \E_{(X,Y) \sim \b P}\bk{C(X,Y)},
    \]
    where the expectation is taken over valid joint distributions of $(X,Y)$
    respecting the marginal distributions. Note that when $\mathcal X = \mathcal
    Y = \R^d$ and $C(x,y) = \norm{x-y}^p$, then the optimal value of the above
    problem for $X,Y$ is the $p$-Wasserstein distance between the marginal
    distributions $\mu_X,\mu_Y$ (up to a power of $1/p$), for discrete,
    $d$-dimensional distributions with finite support.
    \end{rmk}

    It can be shown that the dual problem of the above LP takes the following
    form: \begin{align*}
    \max_{(\b f, \b g) \in \R^{m} \times \R^n} \, & \sum_i a_i f_i + \sum_j b_j
    g_j
    \\ 
    \text{subjected to } & f_i + g_j \le C_{ij}\, \text{ for all $i,j$}.
    \end{align*}
    The dual problem has a neat economic intuition. Suppose Kevin needs to
    transport $\b a$ to $\b b$ but does not understand how. Francisco, his
    profit-maximizing colleague, offers him a deal, where Kevin will pay $f_i$
    for Francisco to
    pick up a unit of mass at $a_i$ and pay $g_j$ for Francisco to dropoff a
    unit of mass at $b_j.$ At minimum, Kevin knows that if any $f_i + g_j > C_
    {ij}$, then Francisco is ripping him off, as transporting one unit from
    $a_i$ to $b_j$ costs exactly $C_{ij}$. If, on the other hand, $f_i + g_j
    \le C_{ij}$ is satisfied, then given any $\b P$, we have \[
    \sum_{i,j} C_{ij}P_{ij} \ge \sum_{i,j} f_i P_{ij} + \sum_{i,j} g_j P_{ij} =
    \sum_i f_i a_i + \sum_j g_j b_j \tag{Weak duality},
    \]
    which means that Kevin cannot lose by taking Francisco's deal. Strong
    duality, which indeed holds, here imply that Francisco's optimal profit is
    zero, and the trade is fair.

    % Note that, fixing $\b f$, the optimal choice of $g_j$ is \[
    % g_j = \min_i (C_{ij} - f_i).
    % \]
    % Let $\b g = \b f^{\b C}$ where $(\b f^{\b C})_j = \min_i (C_{ij} - f_i)$,
    % which is called a $\b C$-transform of $\b f$; we may reformulate the dual
    % problem in terms of only one choice variable $\b f$.

    Given optimal solutions to the primal and dual, $\b P$, $(\b f, \b g)$,
    respectively, \emph{complementary slackness} implies that \[
    P_{ij} (C_{ij} - f_i - g_j) = 0
    \]
    holds, which means that $\b P_{ij} \neq 0 \implies C_{ij} = f_i - g_j$ and
    vice versa.

\end{document}

### 2018-03-24 reports/progress-report/progress.tex
add cite

@@
 \title{Optimal Transport Problems}
-\renewcommand{\b}{\mathbf}
 \newcommand{\R}{\mathbb{R}}
@@
 \newcommand{\E}{\mathbb{E}}
+\usepackage{cleveref}
 \author{Jiafeng Chen\and Francisco Rivera}
 \date{\today}
+\usepackage{hyperref}
+\hypersetup{colorlinks=true, citecolor=Blue, linkcolor=blue}
+\renewcommand{\b}{\mathbf}
 
@@
     \section{Theoretical foundations}
+    The following is a concise review of Chapters 2 and 3 in 
+    \cite{peyre2017computational}.
+
     Let $\b a \in \R^m_{\ge 0}$ and $\b b \in \R^n_{\ge 0}$ be two vectors of
@@
 
+\bibliographystyle{alpha}
+\bibliography{../optimal-transport}
 \end{document}

### 2018-03-27 reports/progress-report/progress.tex
more theoretical results

@@
-\documentclass[11pt]{amsart}
+\documentclass[11pt,reqno]{amsart}
 \usepackage{../am221}
 \linespread{1.2}
+\usepackage[margin=1.5in]{geometry}
 \title{Optimal Transport Problems}
@@
 \newcommand{\E}{\mathbb{E}}
-\usepackage{cleveref}
+%\usepackage[margin=1in]{geometry}
 \author{Jiafeng Chen\and Francisco Rivera}
@@
 \usepackage{hyperref}
-\hypersetup{colorlinks=true, citecolor=Blue, linkcolor=blue}
+\usepackage[capitalise,nameinlink]{cleveref}
+
+\hypersetup{colorlinks=true, citecolor=Blue, linkcolor=Brown}
 \renewcommand{\b}{\mathbf}
 
+\theoremstyle{definition}
+\newtheorem{prop}{Proposition}
 \theoremstyle{remark}
@@
     \section{Theoretical foundations}
-    The following is a concise review of Chapters 2 and 3 in 
+    The following is a concise review of selected topics in Chapters 2 and 3 in 
     \cite{peyre2017computational}.
@@
     respectively, \emph{complementary slackness} implies that \[
-    P_{ij} (C_{ij} - f_i - g_j) = 0
+    P_{ij} (C_{ij} - f_i - g_j) = 0 \tag{Complementary slackness}
     \]
@@
 
+    \subsection{Dual-ascent algorithms and Hungarian algorithm}
+
+    Let $S \subset \{1,\ldots,m\}$ and $S' \subset \{1,\ldots,n\}.$ Let $\one_S$
+    be a vector of zeros with ones at location prescribed by $S$. Given $\b f,
+    \b g$ a feasible pair for the dual, call $(i,j)$ a balanced pair if the
+    condition $f_i + g_j = C_{ij}$
+    binds, and inactive otherwise. Then 
+    \begin{prop}
+
+        Let $(\tilde {\b f}, \tilde {\b g}) = (\b f, \b g) + \epsilon (\one_S,
+        -\one_{S'})$. Then $(\tilde {\b f}, \tilde {\b g})$ is dual feasible for
+        a sufficiently small $\epsilon$ if for all $i$, \[
+        (i,j) \text{ is balanced } \implies \text{$j \in S'$}.
+        \]
+    \label{prop:feasible}
+    \end{prop}
+    \begin{proof}
+        For all $i\in S$, consider \[
+        \epsilon_i = \min_{j: (i,j) \text{ inactive}} C_{ij} - f_i - g_j
+        \]
+        Let $\epsilon = \min_i \epsilon_i$. If $(i,j)$ is inactive, then \[
+        \tilde f_i + \tilde g_j \le f_i + \epsilon + g_j\le C_{ij},
+        \]
+        which continues to be inactive. If $(i,j)$ is balanced, then $\tilde
+        f_i + \tilde g_j = f_i + g_j = C_{ij}$. In either case, $\tilde {\b f},
+        \tilde {\b g}$ continues to be feasible.
+    \end{proof}
+
+    \begin{prop}
+    \label{prop:dualascent}
+        If a dual feasible solution $(\b f, \b g)$ is not optimal, then there
+        exists $S,S'$ such that $(\tilde {\b f}, \tilde {\b g}) = (\b f, \b g) + \epsilon (\one_S,
+        -\one_{S'})$ is feasible for small enough $\epsilon$ with a strictly
+        better objective.
+    \end{prop}
+
+    \begin{proof}
+        Let $\mathcal B$ be the set of balanced edges $(i,j)$, and consider the
+        bipartite graph on $\{1,\ldots,m\} \sqcup \{1,\ldots,n\}$ with edges in
+        $\mathcal B$. Let there be a source node $s$ with capacitated edges to
+        $\{1,\ldots,m\}$ with capacity $a_i$, and a sink node $t$ analogously
+        with capacities $b_j$. Consider the maximum flow $\b F$ on the network
+        via Ford-Fulkerson. If the throughput of the flow is one, then let $P_
+        {ij} = F_{ij}$, which generates a primal feasible solution $\b P$.
+        Strong duality would imply that $\b P, (\b f, \b g)$ are both optimal.
+
+        Thus we may assume that the throughput is less than one. We label the
+        nodes reached from $s$ where $\b F$ does not saturate capacity, as well
+        as the nodes that contribute flow to any labeled nodes. Store the
+        labeled nodes in $Q$. Since the graph is bipartite, $Q$ can be split
+        into $S,S'$. Note that if $(i,j)$ is balanced and $i\in S$, then $j\in
+        S'$. By \cref{prop:feasible}, $(\tilde {\b f}, \tilde {\b g})$ so
+        defined is feasible. We wish to show that \[
+        \one_S^T \b a - \one_{S'}^T \b b > 0.
+        \]
+        We readily conclude that this is indeed the case by accounting for the
+        flow on the network.
+    \end{proof}
+
+    \cref{prop:dualascent} provides a template for the celebrated Hungarian
+    algorithm \cite{kuhn2010hungarian}. The Hungarian algorithm solves the
+    Kantorovich problem in the case when both $\b a, \b b \propto \one$. In a
+    probability context for the Wasserstein distance between empirical measures,
+    if the measures $\mu, \nu$ are both $d$-dimensional with $n$ observations,
+    the Hungarian algorithm is $O(n^3)$ for $d>1$ and $O(n\log n)$ for\footnote{In the $d=1$ case, the optimal transport problem reduces to
+    sorting.} $d =
+    1$
+    \cite{bernton2017inference}.
+
+    \section{Application value}
+
+    \cite{bernton2017inference} is an example of recent work applying the
+    Wasserstein distance to statistics and machine learning. 
+    % Restricting to a subclass of problems called \emph{matching problems}, where
+    % $\b a, \b b \propto \one$.
+
 \bibliographystyle{alpha}

### 2018-03-27 reports/progress-report/progress.tex
Add section stubs to project proposal.

@@
 \newtheorem{rmk}{Remark}
+
 \begin{document}
 \maketitle
+
+\section{Revisions}
+
+By and large, our project goal has remained the same. We have, however,
+focused our scope to focus primarily on the Kantorovich optimal transport
+problem. This is contrast to the more general but intractable continuous
+Wassserstein problem.
+
 \section{Theoretical foundations}
@@
 if the measures $\mu, \nu$ are both $d$-dimensional with $n$ observations,
-    the Hungarian algorithm is $O(n^3)$ for $d>1$ and $O(n\log n)$ for\footnote{In the $d=1$ case, the optimal transport problem reduces to
-    sorting.} $d =
-    1$
-    \cite{bernton2017inference}.
+the Hungarian algorithm is $O(n^3)$ for $d>1$ and $O(n\log n)$
+for\footnote{In the $d=1$ case, the optimal transport problem reduces to
+sorting.} $d = 1$ \cite{bernton2017inference}.
+
+\section{Completed Steps}
+
+% TODO: Francisco will write in code and performance here.
+
+\section{Next Steps}
+
+% TODO: Francisco will write whatever he didn't feel like coding up here.
 
@@
 Wasserstein distance to statistics and machine learning. 
-    % Restricting to a subclass of problems called \emph{matching problems}, where
-    % $\b a, \b b \propto \one$.
+% Restricting to a subclass of problems called \emph{matching problems},
+% where $\b a, \b b \propto \one$.
 

### 2018-03-28 reports/progress-report/progress.tex
comment
Add in code benchmarking to proposal.

--- 2018-03-27 reports/progress-report/progress.tex
@@
 \documentclass[11pt,reqno]{amsart}
+
+% Page set-up
+\usepackage[margin=1.5in]{geometry}
+\linespread{1.2}
+
+% Packages
 \usepackage{../am221}
-\linespread{1.2}
-\usepackage[margin=1.5in]{geometry}
-\title{Optimal Transport Problems}
+\usepackage{bbm}
+\usepackage{hyperref}
+\usepackage[capitalise,nameinlink]{cleveref}
+\hypersetup{colorlinks=true, citecolor=Blue, linkcolor=Brown}
+
+% Commands
+\renewcommand{\b}{\mathbf}
 \newcommand{\R}{\mathbb{R}}
-\usepackage{bbm}
 \newcommand{\one}{\mathbbm{1}}
 \newcommand{\E}{\mathbb{E}}
-%\usepackage[margin=1in]{geometry}
-\author{Jiafeng Chen\and Francisco Rivera}
-\date{\today}
-\usepackage{hyperref}
-\usepackage[capitalise,nameinlink]{cleveref}
 
-\hypersetup{colorlinks=true, citecolor=Blue, linkcolor=Brown}
-\renewcommand{\b}{\mathbf}
-
+% Thm definitions
 \theoremstyle{definition}
@@
 \newtheorem{rmk}{Remark}
+
+\title{Optimal Transport Problems}
+\author{Jiafeng Chen\and Francisco Rivera}
+\date{\today}
+
 \begin{document}
 \maketitle
+
+\section{Revisions}
+
+By and large, our project goal has remained the same. We have, however, focused
+our scope to primarily the Kantorovich optimal transport problem. This is
+contrast to the more general but intractable continuous Wassserstein problem.
+
 \section{Theoretical foundations}
@@
 Let $\b a \in \R^m_{\ge 0}$ and $\b b \in \R^n_{\ge 0}$ be two vectors of
-    nonnegative real numbers such that $\b a^T \one = \b b^T \one = 1$, where
-    $a_i$ and $b_j$ represents the amount of ``mass'' at different positions. 
-    We are
+non-negative real numbers such that $\b a^T \one = \b b^T \one = 1$, where $a_i$
+and $b_j$ represents the amount of ``mass'' at different positions. We are
 interested in a matrix $\b P \in \R^{m \times n}_{\ge 0}$ where $P_{ij}$
-    prescribes
-    the amount of mass flowing from $a_i$ to $b_j$. Naturally, we must have
-    $\sum_j P_{ij} = a_i$ and $\sum_i P_{ij} = b_j$, so that mass is preserved.
-    Let $\b C
-    \in \R^{m\times n}$ be a \emph{cost matrix}, where $C_{ij}$ represents the
-    cost of transporting one unit of $a_i$ to $b_j$. The \emph{Kantorovich
-    optimal transport problem} is the following linear program: \begin{align*}
+prescribes the amount of mass flowing from $a_i$ to $b_j$. Naturally, we must
+have $\sum_j P_{ij} = a_i$ and $\sum_i P_{ij} = b_j$, so that mass is preserved.
+Let $\b C \in \R^{m\times n}$ be a \emph{cost matrix}, where $C_{ij}$ represents
+the cost of transporting one unit of $a_i$ to $b_j$. The \emph{Kantorovich
+optimal transport problem} is the following linear program:
+
+\begin{align*}
 \min_{\b P \in \R^{m \times n}_{\ge 0}} \, & \sum_{i,j} P_{ij} C_{ij} \\
@@
 \begin{rmk}[A probabilistic view]
-    Let $X,Y$ be discrete random variables with support $\b
-    x \in \mathcal X^m$ and $\b y \in \mathcal Y^n$, such that their marginal
-    distributions are
-    prescribed by $P(X = x_i) = a_i$ and $P(Y =
-    y_j) = b_j$. Let $P_{ij} = P(X = x_i, Y = y_j)$, making $\b P = (P_{ij})_
-    {ij}$ a joint
-    distribution. Let $C: \mathcal X \times \mathcal Y \to \R$ be a cost
-    function. Then the
-    Kantorovich problem is equivalent to the following:\[
-    \min_{\b P}\, \E_{(X,Y) \sim \b P}\bk{C(X,Y)},
-    \]
+Let $X,Y$ be discrete random variables with support $\b x \in \mathcal X^m$ and
+$\b y \in \mathcal Y^n$, such that their marginal distributions are prescribed
+by $P(X = x_i) = a_i$ and $P(Y = y_j) = b_j$. Let $P_{ij} = P(X = x_i, Y =
+y_j)$, making $\b P = (P_{ij})_ {ij}$ a joint distribution. Let $C: \mathcal X
+\times \mathcal Y \to \R$ be a cost function. Then the Kantorovich problem is
+equivalent to the following:
+
+\[ \min_{\b P}\, \E_{(X,Y) \sim \b P}\bk{C(X,Y)}, \]
+
 where the expectation is taken over valid joint distributions of $(X,Y)$
-    respecting the marginal distributions. Note that when $\mathcal X = \mathcal
-    Y = \R^d$ and $C(x,y) = \norm{x-y}^p$, then the optimal value of the above
-    problem for $X,Y$ is the $p$-Wasserstein distance between the marginal
-    distributions $\mu_X,\mu_Y$ (up to a power of $1/p$), for discrete,
-    $d$-dimensional distributions with finite support.
+respecting the marginal distributions. Note that when $\mathcal X = \mathcal Y =
+\R^d$ and $C(x,y) = \norm{x-y}^p$, then the optimal value of the above problem
+for $X,Y$ is the $p$-Wasserstein distance between the marginal distributions
+$\mu_X,\mu_Y$ (up to a power of $1/p$), for discrete, $d$-dimensional
+distributions with finite support.
 \end{rmk}
 
-    It can be shown that the dual problem of the above LP takes the following
-    form: \begin{align*}
-    \max_{(\b f, \b g) \in \R^{m} \times \R^n} \, & \sum_i a_i f_i + \sum_j b_j
-    g_j
+It can be shown that the dual problem of the above LP takes the following form:
+
+\begin{align*}
+\max_{(\b f, \b g) \in \R^{m} \times \R^n} \, & \sum_i a_i f_i + \sum_j b_j g_j
 \\ 
@@
 \end{align*}
-    The dual problem has a neat economic intuition. Suppose Kevin needs to
-    transport $\b a$ to $\b b$ but does not understand how. Francisco, his
-    profit-maximizing colleague, offers him a deal, where Kevin will pay $f_i$
-    for Francisco to
-    pick up a unit of mass at $a_i$ and pay $g_j$ for Francisco to dropoff a
-    unit of mass at $b_j.$ At minimum, Kevin knows that if any $f_i + g_j > C_
-    {ij}$, then Francisco is ripping him off, as transporting one unit from
-    $a_i$ to $b_j$ costs exactly $C_{ij}$. If, on the other hand, $f_i + g_j
-    \le C_{ij}$ is satisfied, then given any $\b P$, we have \[
+
+The dual problem has a neat economic intuition. Suppose Kevin needs to transport
+$\b a$ to $\b b$ but does not understand how. Francisco, his profit-maximizing
+colleague, offers him a deal, where Kevin will pay $f_i$ for Francisco to pick
+up a unit of mass at $a_i$ and pay $g_j$ for Francisco to dropoff a unit of mass
+at $b_j.$ At minimum, Kevin knows that if any $f_i + g_j > C_ {ij}$, then
+Francisco is ripping him off, as transporting one unit from $a_i$ to $b_j$ costs
+exactly $C_{ij}$. If, on the other hand, $f_i + g_j \le C_{ij}$ is satisfied,
+then given any $\b P$, we have
+
+\[
 \sum_{i,j} C_{ij}P_{ij} \ge \sum_{i,j} f_i P_{ij} + \sum_{i,j} g_j P_{ij} =
@@
 \]
-    which means that Kevin cannot lose by taking Francisco's deal. Strong
-    duality, which indeed holds, here imply that Francisco's optimal profit is
-    zero, and the trade is fair.
 
-    % Note that, fixing $\b f$, the optimal choice of $g_j$ is \[
-    % g_j = \min_i (C_{ij} - f_i).
-    % \]
+which means that Kevin cannot lose by taking Francisco's deal. Strong duality,
+which indeed holds, here imply that Francisco's optimal profit is zero, and the
+trade is fair.
+
+% Note that, fixing $\b f$, the optimal choice of $g_j$ is
+% \[ g_j = \min_i (C_{ij} - f_i). \]
 % Let $\b g = \b f^{\b C}$ where $(\b f^{\b C})_j = \min_i (C_{ij} - f_i)$,
@@
 Given optimal solutions to the primal and dual, $\b P$, $(\b f, \b g)$,
-    respectively, \emph{complementary slackness} implies that \[
-    P_{ij} (C_{ij} - f_i - g_j) = 0 \tag{Complementary slackness}
-    \]
+respectively, \emph{complementary slackness} implies that
+
+\[ P_{ij} (C_{ij} - f_i - g_j) = 0 \tag{Complementary slackness} \]
+
 holds, which means that $\b P_{ij} \neq 0 \implies C_{ij} = f_i - g_j$ and
@@
 
-    Let $S \subset \{1,\ldots,m\}$ and $S' \subset \{1,\ldots,n\}.$ Let $\one_S$
-    be a vector of zeros with ones at location prescribed by $S$. Given $\b f,
-    \b g$ a feasible pair for the dual, call $(i,j)$ a balanced pair if the
-    condition $f_i + g_j = C_{ij}$
-    binds, and inactive otherwise. Then 
+Let $S \subset \{1,\ldots,m\}$ and $S' \subset \{1,\ldots,n\}.$ Let $\one_S$ be
+a vector of zeros with ones at location prescribed by $S$. Given $\b f, \b g$ a
+feasible pair for the dual, call $(i,j)$ a balanced pair if the condition $f_i +
+g_j = C_{ij}$ binds, and inactive otherwise. Then
+
 \begin{prop}
-
     Let $(\tilde {\b f}, \tilde {\b g}) = (\b f, \b g) + \epsilon (\one_S,
-        -\one_{S'})$. Then $(\tilde {\b f}, \tilde {\b g})$ is dual feasible for
-        a sufficiently small $\epsilon$ if for all $i$, \[
-        (i,j) \text{ is balanced } \implies \text{$j \in S'$}.
-        \]
+    -\one_{S'})$. Then $(\tilde {\b f}, \tilde {\b g})$ is dual feasible for a
+    sufficiently small $\epsilon$ if for all $i$,
+    \[ (i,j) \text{ is balanced } \implies \text{$j \in S'$}. \]
 \label{prop:feasible}
 \end{prop}
+
 \begin{proof}
-        For all $i\in S$, consider \[
-        \epsilon_i = \min_{j: (i,j) \text{ inactive}} C_{ij} - f_i - g_j
-        \]
-        Let $\epsilon = \min_i \epsilon_i$. If $(i,j)$ is inactive, then \[
-        \tilde f_i + \tilde g_j \le f_i + \epsilon + g_j\le C_{ij},
-        \]
-        which continues to be inactive. If $(i,j)$ is balanced, then $\tilde
-        f_i + \tilde g_j = f_i + g_j = C_{ij}$. In either case, $\tilde {\b f},
-        \tilde {\b g}$ continues to be feasible.
+    For all $i\in S$, consider
+    \[ \epsilon_i = \min_{j: (i,j) \text{ inactive}} C_{ij} - f_i - g_j \]
+    Let $\epsilon = \min_i \epsilon_i$. If $(i,j)$ is inactive, then
+    \[ \tilde f_i + \tilde g_j \le f_i + \epsilon + g_j\le C_{ij}, \]
+    which continues to be inactive. If $(i,j)$ is balanced, then $\tilde f_i +
+    \tilde g_j = f_i + g_j = C_{ij}$. In either case, $\tilde {\b f}, \tilde {\b
+    g}$ continues to be feasible.
 \end{proof}
@@
     $\mathcal B$. Let there be a source node $s$ with capacitated edges to
-        $\{1,\ldots,m\}$ with capacity $a_i$, and a sink node $t$ analogously
-        with capacities $b_j$. Consider the maximum flow $\b F$ on the network
-        via Ford-Fulkerson. If the throughput of the flow is one, then let $P_
-        {ij} = F_{ij}$, which generates a primal feasible solution $\b P$.
-        Strong duality would imply that $\b P, (\b f, \b g)$ are both optimal.
+    $\{1,\ldots,m\}$ with capacity $a_i$, and a sink node $t$ analogously with
+    capacities $b_j$. Consider the maximum flow $\b F$ on the network via
+    Ford-Fulkerson. If the throughput of the flow is one, then let $P_ {ij} =
+    F_{ij}$, which generates a primal feasible solution $\b P$. Strong duality
+    would imply that $\b P, (\b f, \b g)$ are both optimal.
 
-        Thus we may assume that the throughput is less than one. We label the
-        nodes reached from $s$ where $\b F$ does not saturate capacity, as well
-        as the nodes that contribute flow to any labeled nodes. Store the
-        labeled nodes in $Q$. Since the graph is bipartite, $Q$ can be split
-        into $S,S'$. Note that if $(i,j)$ is balanced and $i\in S$, then $j\in
-        S'$. By \cref{prop:feasible}, $(\tilde {\b f}, \tilde {\b g})$ so
-        defined is feasible. We wish to show that \[
-        \one_S^T \b a - \one_{S'}^T \b b > 0.
-        \]
+    Thus we may assume that the throughput is less than one. We label the nodes
+    reached from $s$ where $\b F$ does not saturate capacity, as well as the
+    nodes that contribute flow to any labeled nodes. Store the labeled nodes in
+    $Q$. Since the graph is bipartite, $Q$ can be split into $S,S'$. Note that
+    if $(i,j)$ is balanced and $i\in S$, then $j\in S'$. By Proposition
+    \cref{prop:feasible}, $(\tilde {\b f}, \tilde {\b g})$ so defined is
+    feasible. We wish to show that
+
+    \[ \one_S^T \b a - \one_{S'}^T \b b > 0. \]
+
     We readily conclude that this is indeed the case by accounting for the
@@
 Kantorovich problem in the case when both $\b a, \b b \propto \one$. In a
-    probability context for the Wasserstein distance between empirical measures,
-    if the measures $\mu, \nu$ are both $d$-dimensional with $n$ observations,
-    the Hungarian algorithm is $O(n^3)$ for $d>1$ and $O(n\log n)$ for\footnote{In the $d=1$ case, the optimal transport problem reduces to
-    sorting.} $d =
-    1$
+probability context for the Wasserstein distance between empirical measures, if
+the measures $\mu, \nu$ are both $d$-dimensional with $n$ observations, the
+Hungarian algorithm is $O(n^3)$ for $d>1$ and $O(n\log n)$ for\footnote{In the
+$d=1$ case, the optimal transport problem reduces to sorting.} $d = 1$
 \cite{bernton2017inference}.
 
-    \section{Application value}
+\section{Completed Steps}
 
-    \cite{bernton2017inference} is an example of recent work applying the
-    Wasserstein distance to statistics and machine learning. 
-    % Restricting to a subclass of problems called \emph{matching problems}, where
-    % $\b a, \b b \propto \one$.
+We have completed empirical assessments of the LP method as well as the
+Hungarian algorithm through computational experiments. The code which implements
+this can be found on the
+\href{https://github.com/frtennis1/am221-project/tree/master/code}{Github
+repository}.\footnote{\url{https://github.com/frtennis1/am221-project/tree/master/code}}
+
+The Hungarian algorithm's runtime can be visualized, along with a log-log plot
+regression,
+
+\begin{figure}[h]
+\includegraphics[width=\textwidth]{../../code/hungarianruntime}
+\label{fig:hungarianruntime}
+\end{figure}
+
+We can also recreate similar plots for the linear-programming algorithm,
+
+\begin{figure}[h]
+\includegraphics[width=\textwidth]{../../code/lpruntime}
+\end{figure}
+
+\section{Next Steps}
+
+The next steps are two-fold. First of all, we will continue our exploration and
+analysis of existing algorithms to solve the optimal-transport problem. In
+particular, we remain to computationally analyze the dual-ascent algorithms in
+generality, and can also explore the auction algorithm. Secondly, we will
+attempt to approach the problem from a novel direction, with the ultimate
+ambition of arriving at a new algorithm that performs competitively.
+
+% \section{Application value}
+
+% \cite{bernton2017inference} is an example of recent work applying the
+% Wasserstein distance to statistics and machine learning. 
+% Restricting to a subclass of problems called \emph{matching problems},
+% where $\b a, \b b \propto \one$.
 

### 2018-03-28 code/startup.py

@@ -1,0 +1,80 @@
+import numpy as np
+import pandas as pd
+import torch
+import pickle
+
+from tqdm import tqdm_notebook as tqdm
+
+import statsmodels.formula.api as smf
+import statsmodels.api as sm
+import seaborn.apionly as sns
+
+import matplotlib
+import matplotlib.pyplot as plt
+
+plt.style.use('clean')
+sns.set_palette(['#8F223A', '#505B70', '#A6A29F'])
+plt.rcParams['font.sans-serif'] = 'Verdana'
+plt.rcParams['figure.figsize'] = (6.5, 4)
+
+%matplotlib inline
+%config InlineBackend.figure_format = 'retina'
+
+def ft_ax(ax=None,
+          y=1.03,
+          yy=1.1,
+          title=None,
+          subtitle=None,
+          source=None,
+          add_box=False,
+          left_axis=False):
+    """
+    Format either a desired axis or the current axis as an FT plot.
+    """
+
+    if ax is None:
+        ax = plt.gca()
+
+    ax.set_axisbelow(True)
+    
+    if title is not None:
+        title = plt.title(title, y=y, loc='left')
+    if subtitle is not None:
+        plt.annotate(subtitle, xy=title.get_position(),
+                xycoords='axes fraction', xytext=(0,-11), 
+                 textcoords='offset points', size='large') 
+    
+    if source is not None:
+        src = plt.annotate(source, xy=(0,0), 
+             xycoords='axes fraction', xytext=(0,-35), 
+             textcoords='offset points', ha='left', va='top', size='small')
+    
+    # axes and grid-lines
+    plt.grid(axis='y', linewidth=.5)
+    sns.despine(left=True)
+    if not left_axis:
+        ax.yaxis.tick_right()
+        ax.yaxis.set_label_position('right')
+        ax.yaxis.set_label_coords(1,yy)
+        ax.yaxis.get_label().set_rotation(0)
+    ax.tick_params('y', length=0)
+    
+    plt.tight_layout()
+    
+    if add_box:
+        ax2 = plt.axes(ax.get_position().bounds, facecolor=(1,1,1,0))
+        ax2.xaxis.set_visible(False)
+        ax2.yaxis.set_visible(False)
+        x,y = np.array([[.01, 0.15], [y+.12, y+.12]])
+        line = matplotlib.lines.Line2D(x, y, lw=6., color='k')
+        ax2.add_line(line)
+        line.set_clip_on(False)
+    
+    if add_box and source is not None:
[…]

### 2018-04-16 reports/notes-ch4/notes-ch4.tex
notes for ch4

\documentclass[11pt,reqno]{amsart}

% Page set-up
\linespread{1.2}

% Packages
\usepackage{../am221}
\usepackage{bbm}
\usepackage{hyperref}
\usepackage[capitalise,nameinlink]{cleveref}
\hypersetup{colorlinks=true, citecolor=Blue, linkcolor=Brown}

% Commands
\renewcommand{\b}{\mathbf}
\newcommand{\R}{\mathbb{R}}
\newcommand{\one}{\mathbbm{1}}
\newcommand{\E}{\mathbb{E}}
\newcommand{\KL}{\operatorname{KL}}
\newcommand{\Lag}{\operatorname{Lagrange}}

% Thm definitions
\theoremstyle{definition}
\newtheorem{prop}{Proposition}
\theoremstyle{remark}
\newtheorem{rmk}{Remark}

\title{Notes on Entropic Regularization}
\author{Jiafeng Chen\and Francisco Rivera}
\date{\today}

\begin{document}
\maketitle

\section{Statement}

The original problem is \begin{align*}
\min \,& \sum_{i,j} P_{ij} C_{ij}\\
\text{s.t. }&\b P\one = \b a \\
&\b P^T \one = \b b
\end{align*}
\newcommand{\feasible}{\b U(\b a, \b b)}
Denote the feasible region by $\feasible$. \emph{Entropic regularization}
 considers the following modified problem: \[
L^\epsilon_{\b C}(\b a ,\b b) = \min_{\b P \in \feasible} \sum_{i,j} P_{ij}C_
{ij} + \epsilon \underbrace{\sum_
{i,j} P_{ij}
(\log P_{ij} - 1)}_{-H(\b P)}.
\]
We can show that the negative entropy $-H(\b P)$ is $1$-strongly convex (i.e.
 Hessian minus $I$ is
 positive semi-definite). Thus the objective is $\epsilon$-strongly convex. The
 strong convexity means that the optimum $\b P_\epsilon$ is unique.
 Unsurprisingly, $\b P_\epsilon \to \b P^\star$ as $\epsilon \to 0$ where $\b
 P^\star$ is the maximal-entropy solution to the original problem $L_{\b C}^0$.
 Moreover, note that if $\epsilon \to \infty$, we are essentially maximizing the
 entropy, and unsurprisingly, the maximal entropy solution is $\b a \b b^T$:
 In terms of probability, maximal entropy joint distribution is assuming
 independence.

We can reformulate the problem via \emph{KL projection}. Note that minimizing
 the regularized objective is akin to minimizing \begin{align*}
 &\sum_{i,j}\frac{1}{\epsilon}P_{ij}C_{ij} + P_{ij}(\log P_{ij} - 1) + e^{-C_
{ij}/\epsilon} \\
=& \sum_{i,j} P_{ij} \pr{\log P_{ij} - \log\pr{e^{-C_{ij}/\epsilon}}} - P_{ij} +
+ e^{-C_
{ij}/\epsilon} \\
=& \sum_{i,j}P_{ij} \log\frac{P_{ij}}{K_{ij}} - P_{ij} + K_{ij} =: \KL
\binom{\b P}{\b K},
 \end{align*}
where $\b K_{ij} = \exp(-C_{ij}/\epsilon)$ is called a \emph{Gibbs kernel}. Thus
 the problem is akin to projecting $\b K$ onto $\feasible$ via the KL-divergence
 as a distance metric.

The Lagrangian of the regularized problem is \begin{align*}
\Lag(\b P, \b f, \b g) = & \sum_{i,j} P_{ij}C_{ij} + \epsilon \sum_
{i,j} P_{ij}
(\log
 P_{ij} - 1) \\ 
& + \sum_i f_i\pr{a_i - \sum_{j}P_{ij}} + \sum_j g_j \pr{b_j - \sum_i P_{ij}},
\end{align*}
for which the first-order conditions yield \[
C_{ij} + \epsilon \log P_{ij} - f_i - g_j = 0.
\]
Rewriting yields \[
P_{ij} = e^{f_i/\epsilon} e^{-C_{ij} / \epsilon} e^{g_j / \epsilon} = u_i K_{ij}
 v_j
\]
for some $\b u, \b v$. The condition prescribed by $\feasible$, written via $\b
 u, \b v$, yield \[
\b u\odot (\b K \b v) = \b a \quad \b v  \odot(\b K^T \b u) = \b b.
\]
Since the optimum is unique, we need only find $\b u, \b v$ such that these
 conditions hold.

Given $\b v$, we can compute $\b u$ via $\b u = \frac{\b a}{\b K \b v}$ and
 vice versa. This motivates Sinkhorn's algorithm by initializing $\b v^{(0)} =
 \one$ and computing \[
\b u^{(\ell + 1)} = \frac{\b a}{\b K \b v^{(\ell)}} \quad \b v^{(\ell + 1)} = \frac{\b b}
{\b K^T u^{(\ell + 1)}}
\]

It has been shown (2017!) that Sinkhorn updates achieves a $\tau$-approximate
 solution of the unregularized problem in $O(n^2 \log(n)\tau^{-3})$ update
 iterations.

% \bibliographystyle{alpha}
% \bibliography{../optimal-transport}
\end{document}

### 2018-04-16 reports/progress-report/progress.tex
notes for ch4

@@
 % Page set-up
-\usepackage[margin=1.5in]{geometry}
+\usepackage[margin=1in]{geometry}
 \linespread{1.2}
@@
 
+\tableofcontents
+
 \section{Revisions}
@@
 
-\section{Theoretical foundations}
+\section{Completed Steps}
+
+\subsection{Theoretical foundations}
+
 The following is a concise review of selected topics in Chapters 2 and 3 in
@@
 
-\subsection{Dual-ascent algorithms and Hungarian algorithm}
+\subsubsection{Dual-ascent algorithms and Hungarian algorithm}
 
@@
 
-\section{Completed Steps}
+\subsection{Numerical experiments}
 
@@
 The Hungarian algorithm's runtime can be visualized, along with a log-log plot
-regression,
+regression, in \cref{fig:hungarianruntime}. We also include the runtime plots for the simplex algorithm in \cref{fig:hungarianruntime}.
 
@@
 \includegraphics[width=\textwidth]{../../code/hungarianruntime}
+\includegraphics[width=\textwidth]{../../code/lpruntime}
+\caption{Runtime comparison}
 \label{fig:hungarianruntime}
@@
 
-We can also recreate similar plots for the linear-programming algorithm,
-
-\begin{figure}[h]
-\includegraphics[width=\textwidth]{../../code/lpruntime}
-\end{figure}
+% \begin{figure}[h]
+% % \end{figure}
+% %
 
@@
 ambition of arriving at a new algorithm that performs competitively.
+There is also a rich literature and ongoing development on approximate optimization methods, e.g.
+via entropic regularization (Chapter 4 in \cite{peyre2017computational}).
 

### 2018-04-25 reports/paper/paper.tex
paper + code

\documentclass{article}

% if you need to pass options to natbib, use, e.g.:
% \PassOptionsToPackage{numbers, compress}{natbib}
% before loading nips_2018

% ready for submission
\usepackage[preprint]{nips_2018}

% to compile a preprint version, e.g., for submission to arXiv, add
% add the [preprint] option:
% \usepackage[preprint]{nips_2018}

% to compile a camera-ready version, add the [final] option, e.g.:
% \usepackage[final]{nips_2018}

% to avoid loading the natbib package, add option nonatbib:
% \usepackage[nonatbib]{nips_2018}

\usepackage[utf8]{inputenc} % allow utf-8 input
\usepackage[T1]{fontenc}    % use 8-bit T1 fonts
\usepackage{hyperref}       % hyperlinks
\usepackage{url}            % simple URL typesetting
\usepackage{booktabs}       % professional-quality tables
\usepackage{amsfonts}       % blackboard math symbols
\usepackage{nicefrac}       % compact symbols for 1/2, etc.
\usepackage{microtype}      % microtypography
\usepackage{../am221}

% \title{Formatting instructions for NIPS 2018}

% The \author macro works with any number of authors. There are two
% commands used to separate the names and addresses of multiple
% authors: \And and \AND.
%
% Using \And between authors leaves it to LaTeX to determine where to
% break the lines. Using \AND forces a line break at that point. So,
% if LaTeX puts 3 of 4 authors names on the first line, and the last
% on the second line, try using \AND instead of \And before the third
% author name.

% \author{
%   David S.~Hippocampus\thanks{Use footnote for providing further
%     information about author (webpage, alternative
%     address)---\emph{not} for acknowledging funding agencies.} \\
%   Department of Computer Science\\
%   Cranberry-Lemon University\\
%   Pittsburgh, PA 15213 \\
%   \texttt{hippo@cs.cranberry-lemon.edu} \\
%   %% examples of more authors
%   %% \And
%   %% Coauthor \\
%   %% Affiliation \\
%   %% Address \\
%   %% \texttt{email} \\
%   %% \AND
%   %% Coauthor \\
%   %% Affiliation \\
%   %% Address \\
%   %% \texttt{email} \\
%   %% \And
%   %% Coauthor \\
%   %% Affiliation \\
%   %% Address \\
%   %% \texttt{email} \\
%   %% \And
%   %% Coauthor \\
%   %% Affiliation \\
%   %% Address \\
%   %% \texttt{email} \\
% }

% Commands
\renewcommand{\b}{\mathbf}
\newcommand{\R}{\mathbb{R}}
\newcommand{\one}{\mathbbm{1}}
\newcommand{\E}{\mathbb{E}}

% Thm definitions
\theoremstyle{definition}
\newtheorem{prop}{Proposition}
\theoremstyle{remark}
\newtheorem{rmk}{Remark}

\title{420}
\author{Jiafeng Chen\\School of Engineering and Applied Sciences\\Harvard University\\\url{jiafengchen@college.harvard.edu}\And Francisco Rivera\\School of Engineering and Applied Sciences\\Harvard University\\\url{frivera@college.harvard.edu}}
\date{\today}

\begin{document}
\maketitle

\section{Introduction and Background}

\subsection{The Optimal Transport Problem}

\subsection{Applications} % (fold)
\label{sub:applications}

% subsection applications (end)

\section{Exact Methods}

\subsection{Linear Programming} % (fold)
\label{sub:linear_programming}

% subsection linear_programming (end)

\subsection{Hungarian Algorithm} % (fold)
\label{sub:hungarian_algorithm}

% subsection hungarian_algorithm (end)

\section{Inexact Methods} % (fold)
\label{sec:inexact_methods}

\subsection{Entropic Regularization and Sinkhorn's Algorithm} % (fold)
\label{sub:entropic_regularization_and_sinkhorn_s_algorithm}

\subsubsection{Convergence}

\subsubsection{Numerical Experiments}

% subsection entropic_regularization_and_sinkhorn_s_algorithm (end)

\subsection{Genetic Algorithms (Hokey)} % (fold)
\label{sub:genetic_algorithms_}

% subsection genetic_algorithms_ (end)
% section inexact_methods (end)

\section{Applications and Experiments} % (fold)
\label{sec:applications_}

% section applications_ (end)

\section{Conclusion} % (fold)
\label{sec:conclusion}

% section conclusion (end)

\bibliographystyle{alpha}
\bibliography{../optimal-transport}
\end{document}

### 2018-04-25 reports/paper/paper.tex
50% of paper

@@
 \usepackage{hyperref}       % hyperlinks
+\hypersetup{colorlinks=true, citecolor=Blue, linkcolor=Brown}
+
+\usepackage[capitalise,nameinlink]{cleveref}
+
 \usepackage{url}            % simple URL typesetting
@@
 \usepackage{../am221}
+\usepackage{algpseudocode,algorithm,algorithmicx}
+\usepackage{bbm}
+\usepackage{amsthm}
 
+\newtheorem{theorem}{Theorem}
+
+\linespread{1.1}
 % \title{Formatting instructions for NIPS 2018}
@@
 
+\newcommand{\KL}{\operatorname{KL}}
 % Thm definitions
@@
 
-\title{420}
+\title{Let's Get Movin': Frontiers of Optimal Transport}
 \author{Jiafeng Chen\\School of Engineering and Applied Sciences\\Harvard University\\\url{jiafengchen@college.harvard.edu}\And Francisco Rivera\\School of Engineering and Applied Sciences\\Harvard University\\\url{frivera@college.harvard.edu}}
@@
 
+\newcommand{\feasible}{\b U(\b a, \b b)}
+
 \begin{document}
 \maketitle
-
 \section{Introduction and Background}
 
-\subsection{The Optimal Transport Problem}
+\subsection{The Optimal Transport Problem and Basic Properties}
 
+Let $\b a \in \R^n_{\ge 0}$ and $\b b \in \R^m_{\ge 0}$ be two vectors of
+non-negative real numbers such that $\b a^T \one = \b b^T \one = 1$, where $a_i$
+and $b_j$ represents the amount of ``mass'' at different positions. We are
+interested in a matrix $\b P \in \R^{n\times m}_{\ge 0}$ where $P_{ij}$
+prescribes the amount of mass flowing from $a_i$ to $b_j$. Naturally, we must
+have $\sum_j P_{ij} = a_i$ and $\sum_i P_{ij} = b_j$, so that mass is preserved.
+Let $\b C \in \R^{n\times m}$ be a \emph{cost matrix}, where $C_{ij}$ represents
+the cost of transporting one unit of $a_i$ to $b_j$. The \emph{Kantorovich
+optimal transport problem} is the following linear program:
+\begin{equation}
+\label{eq:ot_problem}
+\min_{\b P \in \feasible} \, \sum_{i,j} P_{ij} C_{ij} \tag{Kantorovich}
+\end{equation}
+where \[
+\feasible = \{\b P \in \R^{n\times m}_{\ge 0}: \b P \one = \b a, \b P^T \one =
+ \b b\}.
+\]
+
+\begin{rmk}[A probabilistic view]
+Let $X,Y$ be discrete random variables with support $\b x \in \mathcal X^m$ and
+$\b y \in \mathcal Y^n$, such that their marginal distributions are prescribed
+by $P(X = x_i) = a_i$ and $P(Y = y_j) = b_j$. Let $P_{ij} = P(X = x_i, Y =
+y_j)$, making $\b P = (P_{ij})_ {ij}$ a joint distribution. Let $C: \mathcal X
+\times \mathcal Y \to \R$ be a cost function. Then the \eqref{eq:ot_problem} is
+equivalent to the following:
+
+\[ \min_{\b P}\, \E_{(X,Y) \sim \b P}\bk{C(X,Y)}, \]
+
+where the expectation is taken over valid joint distributions of $(X,Y)$
+respecting the marginal distributions. Note that when $\mathcal X = \mathcal Y =
+\R^d$ and $C(x,y) = \norm{x-y}^p$, then the optimal value of the above problem
+for $X,Y$ is the $p$-Wasserstein distance between the marginal distributions
+$\mu_X,\mu_Y$ (up to a power of $1/p$), for discrete, $d$-dimensional
+distributions with finite support.
+\end{rmk}
+
+\paragraph{Duality} It can be shown that the dual problem of the above LP takes
+ the following form:
+
+\begin{align*}
+\max_{(\b f, \b g) \in \R^{m} \times \R^n} \, & \sum_i a_i f_i + \sum_j b_j g_j
+\\ 
+\text{subjected to } & f_i + g_j \le C_{ij}\, \text{ for all $i,j$}.
+\end{align*}
+
+The dual problem has a neat economic intuition. Suppose Kevin needs to transport
+$\b a$ to $\b b$ but does not understand how. Francisco, his profit-maximizing
+colleague, offers him a deal, where Kevin will pay $f_i$ for Francisco to pick
+up a unit of mass at $a_i$ and pay $g_j$ for Francisco to dropoff a unit of mass
+at $b_j.$ At minimum, Kevin knows that if any $f_i + g_j > C_ {ij}$, then
+Francisco is ripping him off, as transporting one unit from $a_i$ to $b_j$ costs
+exactly $C_{ij}$. If, on the other hand, $f_i + g_j \le C_{ij}$ is satisfied,
+then given any $\b P$, we have
+
+\[
+\sum_{i,j} C_{ij}P_{ij} \ge \sum_{i,j} f_i P_{ij} + \sum_{i,j} g_j P_{ij} =
+\sum_i f_i a_i + \sum_j g_j b_j \tag{Weak duality},
+\]
+
+which means that Kevin cannot lose by taking Francisco's deal. Strong duality,
+which indeed holds, here imply that Francisco's optimal profit is zero, and the
+trade is fair.
+
+% Note that, fixing $\b f$, the optimal choice of $g_j$ is
+% \[ g_j = \min_i (C_{ij} - f_i). \]
+% Let $\b g = \b f^{\b C}$ where $(\b f^{\b C})_j = \min_i (C_{ij} - f_i)$,
+% which is called a $\b C$-transform of $\b f$; we may reformulate the dual
+% problem in terms of only one choice variable $\b f$.
+
+Given optimal solutions to the primal and dual, $\b P$, $(\b f, \b g)$,
+respectively, \emph{complementary slackness} implies that
+
+\[ P_{ij} (C_{ij} - f_i - g_j) = 0 \tag{Complementary slackness} \]
+
+holds, which means that $\b P_{ij} \neq 0 \implies C_{ij} = f_i - g_j$ and
+vice versa. 
 \subsection{Applications} % (fold)
@@
 
+Let \[H(\b P)= -\sum_{i,j} P_{ij}(\log P_{ij} - 1)\] be the \emph{entropy} of
+ the transport matrix $\b P$. Observe that \[
+\nabla^2_{\b P} (-H(\b P)) - I \text{ is positive semidefinite},
+\]
+and so $-H(\b P)$ is 1-strongly convex. We regularize the objective of 
+\eqref{eq:ot_problem} by $-\epsilon H(\b P):$ \begin{equation}
+\label{eq:entropic}
+  \min_{\b P \in \feasible} \sum_{i,j}P_{ij}C_{ij} - \epsilon H(\b P) 
+\tag{Entropic Regularization}
+\end{equation}
+so as to make the objective $\epsilon$-strongly convex. \emph{Entropic
+ regularization} approximates \eqref{eq:ot_problem} by solving 
+\eqref{eq:entropic}, where \[
+\argmin_{\b P \in \feasible} \sum_{i,j}P_{ij}C_{ij} - \epsilon H(\b P) 
+ {\underset{n\to\infty}{\longrightarrow}}
+ \argmin_{\b P \in \feasible} \sum_{i,j}P_{ij}C_{ij},
+\]
+where the convergence is towards the maximal-entropy solution of 
+\eqref{eq:ot_problem}.
+
+Leveraging duality theory, the Lagrangian of the regularized problem is 
+\begin{align*}
+\mathcal L(\b P, \b f, \b g) = & \sum_{i,j} P_{ij}C_{ij} + \epsilon \sum_
+{i,j} P_{ij}
+(\log
+ P_{ij} - 1) \\ 
+& + \sum_i f_i\pr{a_i - \sum_{j}P_{ij}} + \sum_j g_j \pr{b_j - \sum_i P_{ij}},
+\end{align*}
+where the first-order condition on $\b P$ shows that the optimum $\b P^\*$
+ satisfies 
+\begin{equation}
+\label{eq:foc}
+  C_{ij} + \epsilon \log P_{ij}^\* - f_i^\star - g_j^\* = 0 \implies \b P^\* =
+ \diag(\exp(\b f^\* / \epsilon)) \exp(-\b C/\epsilon )\diag(\exp(\b g^\* /
+ \epsilon))
+\end{equation}
+for some $\b f^\star, \b g^\*,$ where the $\exp$ is taken component-wise.
+ Suppose $\b f^\star, \b g^\*$ is such that $\b P^\star$ defined by 
+\eqref{eq:foc} is within the constraints $\feasible$, then the value of the
+ Lagrangian $\mathcal L(\b P^\*, \b f^\*, \b g^\*)$ equals the value of 
+\eqref{eq:entropic}, and since  $\b P^\*$ is chosen to maximize the strongly
+ concave function $\b P \mapsto \mathcal L(\b P, \b f^\*, \b g^\*)$, we have
+ that $(\b P^\*, \b f^\*, \b g^\*)$ solves both programs.
+
+Before proceeding, we define some notation that is a hybrid of 
+\cite{peyre2017computational} and \cite{altschuler2017near}. Let $\b A = \exp
+(-\b C / \epsilon),$ $\b u = \exp(\b f / \epsilon)$, $\b v = \exp(\b g /
+ \epsilon)$, $\b x = \log \b u$, $\b y = \log \b v$, $\b X = \diag(\b u)$, $\b
+ Y = \diag(\b v)$, where $\exp, \log$ are taken componentwise. Thus, we aim to
+ find vectors $\b u, \b v$, such that \[
+\b P^\* = \b X \b A \b Y.
+\]
+satisfies the constraints $\feasible.$ In terms of $\b u, \b v$, the constraints
+ are \begin{equation}
+\label{eq:sinkhorn}
+   \b u \odot (\b A \b v) = \b a \quad \text{    and    } \quad \b v \odot (\b A^T
+ \b u) = \b b 
+ \end{equation}
+\eqref{eq:sinkhorn} shows that, given $\b v$, we can only set $\b u = \frac{\b
+ a }{\b A \b v}$, where division is componentwise. This motivates the
+ Sinkhorn-Knopp algorithm \cite{sinkhorn1967concerning}, which iteratively
+ updates
+ $\b u, \b v$ such that one of the constraints hold exactly in this fashion. In
+ practical implementation, to prevent numerical underflow, the algorithm should
+ work in log-space with $\b x, \b y$. \cref{alg:sinkhorn} presents a numerically
+ stable implementation of Sinkhorn-Knopp that terminates when the $\ell_1$ error
+ on constraint violation gets sufficiently small. At each step of 
+\cref{alg:sinkhorn}, if $k$ is odd, then $\b P^{(k)}$ matches row-wise with $\b
+ a$, and if $k$ is even, then $\b P^{(k)}$ matches column-wise with $\b
+ b$.
+
+\begin{algorithm}
+\caption{Numerically stable Sinkhorn-Knopp algorithm as presented in 
+\cite{altschuler2017near}}
+\label{alg:sinkhorn}
+
+\begin{algorithmic}
+
+\Function{Sinkhorn}{$\b A, \b a, \b b, \epsilon$}
+\State Initialize $k=0$
+\State $\b P^{(0)} \gets \b A, \b x^0 \gets 0, \b y^0 \gets 0$
+\While{$\norm{\b P^{(k)}\one - \b a}_1 + \norm{(\b P^{(k)})^T\one - \b b}_1 >
+ \epsilon$}
+\State $k\gets k + 1$
+\If{$k$ is odd}
+\State $\Delta \b x \gets \log\frac{\b a}{\b P^{(k-1)}\one}$
+\State $\b x^k \gets \b x^{k-1} + \Delta \b x, \b y^k \gets \b y^{k-1}$
+\EndIf
+\If{$k$ is even}
+\State $\Delta \b y \gets \log\frac{\b b}{(\b P^{(k-1)})^T\one}$
+\State $\b x^k \gets \b x^{k-1}, \b y^k \gets \b y^{k-1}  + \Delta \b y$
+\EndIf
+\State $\b P^{(k)} \gets \b X^k \b A \b Y^{k}$
+\EndWhile
+\State\Return{$\b P^{(k)} / \norm{\b P^{(k)}}_1$}
+\Comment{$\b P^{(0)}$ is not normalized; $\b P^{(k)}, k > 0$ is normalized.}
+\EndFunction
+\end{algorithmic}
+\end{algorithm}
+
+Note that \cref{alg:sinkhorn} does not output a matrix that satisfies the
+ conditions $\feasible$ exactly. However, \cite[Algorithm 2]{altschuler2017near}
+ analyzes a rounding scheme that forces the output to be inside $\feasible$. The
+ rounding scheme introduces error on the order of $\epsilon$ and thus does not
+ pose any problems for solving the original program \eqref{eq:entropic}. The
+ rest of this section will be devoted to an analysis of \cref{alg:sinkhorn}.
+
 \subsubsection{Convergence}
+\emph{A priori}, it is unclear that \cref{alg:sinkhorn} converges, nor is the
+ speed of convergence obvious. 
+\cite{altschuler2017near} presents an extremely succinct analysis of 
+\cref{alg:sinkhorn}, which we reproduce in this section. Let \[
+f(\b x, \b y) = \sum_{i,j} A_{ij} e^{x_i + y_j} - \sum_i a_i x_i - \sum_j b_j
+ y_j.
+\] 
+The first order conditions for the optimality of $f$ gives $\b x, \b y$
+ that ensures \eqref{eq:sinkhorn}. Sinkhorn-Knopp can be viewed as coordinate
+ descent on $f$: To optimize $x_i^k$, the first order condition yields \[
+x_i^k = \log \frac{a_i}{\sum_j A_{ij} e^{y_j^{k-1}}} = \log \frac{a_i}{e^
+{x_i^{k-1}}\sum_j A_{ij} e^{y_j^{k-1}}} + x_i^{k-1} = \log \frac{a_i}{\b P^{
+(k-1)}\one} + x_i^{k-1},
+\]
+which is exactly the update rule in \cref{alg:sinkhorn}.
+
+The proof of \cite{altschuler2017near} uses the follow strategy. First, we
+ calculate the step gain in the Sinkhorn-Knopp algorithm for optimizing $f$ in
+ terms of Kullback-Leibler divergences.
+ Next we bound the distance of $f$'s initial value from its optimal value.
+ Lastly, using Pinsker's inequality, we show that if \cref{alg:sinkhorn} does
+ not terminate, then the step size must be larger than $C\epsilon^{2}$, which
+ then gives a bound of convergence in $\epsilon^{-2}$.
+
+For $k > 1$, we can
+ calculate the gain of each step \cite[Lemma 2]{altschuler2017near}:
+\begin{align}
+f(\b x^{k-1}, \b y^{k-1}) - f(\b x^{k}, \b y^{k}) &= \pr{\sum_{i,j} A_{ij}e^
+{x_i^{k-1} + y_i^{k-1}} - \sum_{i,j} A_{ij}e^
+{x_i^{k} + y_i^{k}}} \nonumber\\&\qquad+ \ip{\b a, \b x^k -\b x^{k-1}} + \ip{\b
+ b,
+ \b y^k - \b y^{k-1}}
+ \nonumber\\
+&= (1-1) + \KL(\b a || \b P^{(k)}\one) + \KL(\b b || (\b P^{(k)})^T\one)
+ \nonumber \tag{Writing out the inner products}\\
+&= \KL(\b a || \b P^{(k)}\one) + \KL(\b b || (\b P^{(k)})^T\one). 
+\label{eq:klsum}
+\end{align}
+\eqref{eq:klsum} gives an elegant formulation of the gain each step exactly as a
+ function of the Kullback-Leibler divergence between the target distribution and
+ the current marginal distribution.
+
+Next, we claim that the total distance in $f$ that we are required to
+ travel is bounded above \cite[Lemma 3]{altschuler2017near}. Note that 
+\cref{alg:sinkhorn} behaves almost identically---producing the same $\b P^{(k)}$
+ at each iteration---if the input $\b A$ passed in is
+ multiplied by a constant. Thus we may without loss of generality replace $\b A$
+ with
+ $\b A_0 = \b A / \norm{\b A}_1$. Let $f^k$ denote $f(\b x^k, \b y^k)$ and
+ $f^\*$ denote $\min f$. Then we have $
+f(0,0) - f^1 = \KL(\b a || \b A_0\one) \ge 0.
+$
+Thus \[
+f^1 - f^\* \le f(0,0) - f^\* = \sum_{i} a_i x_i^\* + \sum_j b_j y_j^\* \le \max_
+{i,j}
+(x_i^\* + y_j^\*).
+\]
+Observe that $
+A_{0, ij}e^{x_i^\* + y_j^\*} \le \sum  A_{0, ij}e^{x_i^\* + y_j^\*} = 1.
+$
+Therefore $x_i^\* + y_j^\* \le -\log A_{0,ij} \le \log \frac{\norm{\b A}_1}{\min_
+{i,j} A_{ij}}$. Putting these together shows that \begin{equation}
+\label{eq:total}
+  f^1 - f^\* \le \log \frac{\norm{\b A}_1}{\min_
+{i,j} A_{ij}}.
+\end{equation}
+
+Lastly, Pinsker's inequality \cite[Lemma 4]{altschuler2017near}, a well-known
+ result in information theory, states that for any probability measures $p, q$,
+ \[
+\norm{p-q}_1^2 \le 2 \KL(p||q). 
+\]
+Applying Pinsker's inequality shows that every step in \cref{alg:sinkhorn}
+ before termination has \[
+\epsilon < \norm{\b P^{(k)}\one - \b a}_1 + \norm{(\b P^{(k)})^T\one - \b b}_1
+ \le \bk{4\pr{\KL(\b a || \b P^{(k)}\one) + \KL(\b b || (\b P^{(k)})^T\one)}}^
+{1/2}.
+\]
+Therefore we improve $f$ by more than $\frac{1}{4}\epsilon^2$ every step before
+ termination. Since the total improvement is bounded by \eqref{eq:total}, we
+ arrive at the main result of \cite{altschuler2017near}:
+\begin{theorem}[\cite{altschuler2017near}]
+\label{thm:sinkhorn_convergence}
+  \cref{alg:sinkhorn} terminates in at most $4\epsilon^{-2}\log \frac{\norm{\b A}_1}{\min_
+{i,j} A_{ij}}$ steps.
+\end{theorem}
+
+\begin{figure}[tb]
+  \centering
+  \includegraphics{figs/sinkhorn.eps}
+  \caption{Numerical experiments on the speed of $\ell_1$ convergence for \cref{alg:sinkhorn}}
+  \label{fig:sinkhorn_convergence}
+\end{figure}
 
 \subsubsection{Numerical Experiments}
+\label{sub:numerical_sink}
+We implement \cref{alg:sinkhorn} and compute the $\ell_1$ distance $\norm{\b P^{
+(k)}\one - \b a}_1 + \norm{(\b P^{(k)})^T\one - \b b}_1$ as a function of $k$.
+ We generate random marginal distribution targets $\b a, \b b$ by sampling
+ from a uniform distribution on the unit interval and normalizing by the sum. We
+ generate random cost matrices by sampling each component from a uniform
+ distribution. We plot the log-log plot in \cref{fig:sinkhorn_convergence}. Note
+ that if the results of \cref{thm:sinkhorn_convergence} is tight for these
+ random matrices and constraints, then the log-log plot should look linear with
+ slope $-2$. This is not what we observe in \cref{fig:sinkhorn_convergence},
+ suggesting that the average case---if not worst case---of \cref{alg:sinkhorn}
+ may be better than $O(\epsilon^{-2})$. In fact, if we plot the log of error
+ against $k$, we find that the relationship is linear, suggesting that, for these
+ random matrices and distributions, the behavior of \cref{alg:sinkhorn} is
+ $O(\log(1/\epsilon))$.
 
 % subsection entropic_regularization_and_sinkhorn_s_algorithm (end)
+
+\subsubsection{Towards a Tightness Analysis of \cref{thm:sinkhorn_convergence}}
+\cref{sub:numerical_sink} suggests that the bound of $\epsilon^{-2}$ may be far
+ too generous. In this section we sketch a potential direction in analyzing the
+ worst-case performance of \cref{alg:sinkhorn}. \cite{sason2015reverse} gives a
+ reverse Pinsker's inequality: For discrete probability measures on finite
+ support $p,q$, \[
+\KL(p||q) \le \log \pr{1+\frac{1}{q_{\min}}\norm{p-q}_1^2} \le \frac{1}{q_{\min}}\norm{p-q}_1^2,
+\]
+where $q_{\min}$ is the minimum of $q$ on its support. The inequality suggests
+ that before termination and after the $\ell_1$ error reaches some $\delta$,
+ every
+ step of \cref{alg:sinkhorn} improves $f$ by at
+ most \[
+\KL(\b a || \b P^{(k)}\one) + \KL(\b b || (\b P^{(k)})^T\one) \le \frac{C}
+{\min(\b P^{(k)}\one, (\b P^{(k)})^T\one)}\delta^2,
+\]
+for some constant $C$, where the minimum is taken element-wise. We can remove
+ the dependence on $k$ of
+ the right-hand side by the following consideration. Let $k'$ be the first time
+ that the $\ell_1$-error is below $\frac{1}{2}\min(\b a, \b b)$. Then for every
+ $k > k'$, $\min(\b P^{(k)}\one, (\b P^{(k)})^T\one) \ge \frac{1}{2}\min(\b a,
+ \b b)$, and so the bound becomes \begin{equation}
+   \label{eq:bound_improvement}
+\KL(\b a || \b P^{(k)}\one) + \KL(\b b || (\b P^{(k)})^T\one) \le \frac{C'}
+{\min(\b a,
+ \b b)}\delta^2,
+ \end{equation}
+for $k > k'$. Let $k_\epsilon$ be the step immediately before termination. Let
+ $k_\delta$ be the first iteration such that $k_\delta > k'$ and that $k_\delta$
+ reaches $\ell_1$-error of $\delta$.
+If we
+ can show that there is some inputs $\b A, \b a, \b
+ b,\epsilon$ for
+ which we can bound the total improvement $f^{k_\eta} - f^{k_\epsilon}$ below by
+ some quantity $M(\b A, \b a, \b
+ b,\epsilon, \delta)$, then for the inputs $\b A, \b a, \b
+ b,\epsilon$, \cref{alg:sinkhorn} has a lower bound of $\Omega\pr{M(\b A, \b a,
+ \b
+ b,\epsilon, \delta)\delta^{-2}\min(\b a, \b b)}$. Here, $\delta$ should be
+ implicitly a function of $\epsilon$, e.g. $\delta = \sqrt{\epsilon}$, so that
+ the bound is a function of $\epsilon$. Choosing $\delta(\epsilon)$ such that
+ the bound $M(\b A, \b a,
+ \b
+ b,\epsilon, \delta)$ is tractable is a major difficulty, which we leave to
+ future work.
 

### 2018-04-25 reports/paper/paper.tex
more on paper; genetic section

@@
 where the convergence is towards the maximal-entropy solution of 
-\eqref{eq:ot_problem}.
+\eqref{eq:ot_problem}. The convergence property makes \eqref{eq:entropic} an
+ attractive alternative to solving optimal transport; other reasons why 
+\eqref{eq:entropic} is useful is explored in a seminal article by 
+\cite{cuturi2013sinkhorn}.
 
@@
 which is exactly the update rule in \cref{alg:sinkhorn}.
+
+\begin{rmk}
+  We do not need to limit ourselves to using coordinate descent to minimize $f$.
+ For instance
+\cite{brauer2017sinkhorn} uses Newton's method, and achieves similar convergence
+ performance as \cref{alg:sinkhorn}.
+\end{rmk}
 
@@
 
-\subsection{Genetic Algorithms (Hokey)} % (fold)
+\subsection{Genetic Algorithm} % (fold)
 \label{sub:genetic_algorithms_}
+
+Recent advances in machine learning points towards the versatility of genetic
+ algorithms, even as a competitive alternative to deep learning based on
+ gradient methods \cite{such2017deep}. Inspired by these advances, we consider
+ the following genetic algorithm
+
+\begin{algorithm}
+\caption{A toy genetic algorithm for optimal transport}
+\label{alg:genetic}
+
+  \begin{algorithmic}
+\Function{GeneticOT}{$\phi, \gamma, k, \b a, \b b, T$}
+\State $\mathsf{initial} \gets \b a \b b^T$
+\State $\mathsf{family} \gets \mathsf{Array}[\phi]$, initialized to $
+\mathsf{initial}$
+\State $\mathsf{children} \gets \mathsf{Array}[\phi\gamma]$, initialized to
+ empty
+  \For{$\mathsf{generation}$ in $1:T$}
+    \For{$\mathsf{person}$ in $\mathsf{family, \gamma}$}
+      \For{$\mathsf{person}'$ in \Call{Subset}{$\mathsf{family}$}} \Comment
+{\textsc{Subset} chooses a subset of size $\gamma$}
+\State $\mathsf{child} \gets$ \Call{Marry}{$\mathsf{person}, 
+\mathsf{person}'$}
+\State $\mathsf{child} \gets$ \Call{Perturb}{$\mathsf{child},k$}
+\State Add $\mathsf{child}$ to $\mathsf{children}$
+\EndFor
+\EndFor
+\State Choose the $\phi$ elements in $\mathsf{children}$ with minimal cost for
+ \eqref{eq:ot_problem} and set to $\mathsf{family}$
+\EndFor
+\State\Return{element in $\mathsf{family}$ that achieves minimal cost}
+\EndFunction
+  \end{algorithmic}
+\end{algorithm}
+Here \textsc{Marry} picks two probability matrices and averages them
+ elementwise, and \textsc{Perturb} chooses random indices $i_1,i_2, j_1,j_2$ and
+ moves mass in the probability matrix in such a way that preserves the marginal
+ distributions and respects the constraint that $0\le p \le 1$ for all
+ components $p$. Such perturbations are repeated $k$ times. $\phi$ is a
+ parameter that controls the size of families, and $\gamma$ is a parameter that
+ decides the number of offsprings. At each generation, we take the top $\phi$
+ matrices that achieve the lowest cost. We hoped that the population drifts
+ towards the optimal transport matrix as the number of generation increases.
+
+The empirical results we obtain are less than encouraging, however. We plot the
+ optimal cost each generation with $n = m = 4, \phi=100, \gamma =5$ for various
+ values of $k$ in \cref{fig:genetic_fail}. Curiously, the improvement in the
+ first generation is always large, with a magnitude of improvement increasing
+ in $k$, but the algorithm seems to achieve some non-optimal steady state as
+ generation goes on. Of course, \cref{alg:genetic} can potentially be improved
+ by thinking about the design of \textsc{Marry} and \textsc{Perturb} more
+ carefully, perhaps leveraging some information about $\b C$.
+
+\begin{figure}[tb]
+  \centering
+  \includegraphics{figs/genetic.eps}
+  \caption{Non-convergence of \cref{alg:genetic}}
+  \label{fig:genetic_fail}
+\end{figure}
 

### 2018-04-26 reports/paper/paper.tex
Add LP section to paper

@@
 \label{sub:linear_programming}
+
+The \ref{eq:ot_problem} Optimal Transport problem as formulated is an LP. This
+means that a natural first-step to benchmarking the problem difficulty is to use
+a generic LP solver. The only difficulty we face when doing this is that we must
+flatten our arguments from matrices to vectors in order to fit the API of a
+generic LP solver, such as \texttt{cvxopt.solvers.lp} which we employ. Recall
+that our problem inputs are $\mathbf{a} \in \mathbb{R}^n_{\ge 0}$, $\mathbf{b}
+\in \mathbb{R}^m_{\ge 0}$ with sum of entries equal to 1 and $\mathbf{C} \in
+\mathbb{R}^{n,m}$. We flatten $C$ into a vector $\mathbf{c} \in \mathbb{R}^{nm}$
+and define the matrix
+\[ \mathbf{A} := \begin{bmatrix} I_n \otimes \mathbf{1}_m \\ \mathbf{1}_n
+\otimes I_m \end{bmatrix} \]
+which allows us to formulate our LP as,
+\[ \min \mathbf{c}^\top \mathbf{x} \text{ such that } \mathbf{x} \ge 0, \mathbf{A} \mathbf{x} = \begin{bmatrix}
+\mathbf{a} \\ \mathbf{b} \end{bmatrix}.\]
+
+We may then benchmark the performance of this data by generating data. We
+generate random square $n \times n$ cost matrices by making each entry i.i.d.
+$\chi^2(1)$, and perform optimal transport on $\mathbf{a} = \mathbf{b} =
+\mathbf{1}_n / n$. The runtimes for different values of $n$ are documented in
+\cref{fig:lpruntime}.
+
+\begin{figure}[h]
+\includegraphics[width=\textwidth]{../../code/lpruntime}
+\caption{Runtime comparison}
+\label{fig:lpruntime}
+\end{figure}
 
@@
 % subsection genetic_algorithms_ (end)
-% section inexact_methods (end)
 
@@
 
-% section applications_ (end)
-
 \section{Conclusion} % (fold)
 \label{sec:conclusion}
-
-% section conclusion (end)
 

### 2018-04-26 reports/paper/paper.tex
.
Add MNIST section to paper

@@
 which allows us to formulate our LP as,
-\[ \min \mathbf{c}^\top \mathbf{x} \text{ such that } \mathbf{x} \ge 0, \mathbf{A} \mathbf{x} = \begin{bmatrix}
-\mathbf{a} \\ \mathbf{b} \end{bmatrix}.\]
+\[ \min \mathbf{c}^\top \mathbf{x} \text{ such that } \mathbf{x} \ge 0,
+\mathbf{A} \mathbf{x} = \begin{bmatrix} \mathbf{a} \\ \mathbf{b}
+\end{bmatrix}.\]
+
+\begin{figure}[b]
+\includegraphics[width=\textwidth]{../../code/lpruntime}
+\caption{LP Solver Runtime}
+\label{fig:lpruntime}
+\end{figure}
 
@@
 \mathbf{1}_n / n$. The runtimes for different values of $n$ are documented in
-\cref{fig:lpruntime}.
-
-\begin{figure}[h]
-\includegraphics[width=\textwidth]{../../code/lpruntime}
-\caption{Runtime comparison}
-\label{fig:lpruntime}
-\end{figure}
+\cref{fig:lpruntime}. While this gives us a baseline to how quickly we can
+compute optimal transport problems, solving the problem for $n$ bigger than a
+couple dozen quickly becomes an expensive task. Thus, we may look elsewhere for
+more efficient solutions that either solve a special case or provide an
+approximate solution.
 
@@
 \label{sub:hungarian_algorithm}
+
+One such special case of the optimal transport problem is when $\mathbf{a} =
+\mathbf{b} \propto \mathbf{1}_n$ as in our simulations to test the performance
+of the generic LP solver. In this case, mass is being matched in unit pieces,
+which makes the problem more tractable. The Hungarian algorithm solves this
+special-case.
+
+% TODO: exposition on how Hungarian works
+
+\begin{figure}[h]
+\includegraphics[width=\textwidth]{../../code/hungarianruntime.pdf}
+\caption{Hungariant Algorithm runtime}
+\label{fig:hungarianruntime}
+\end{figure}
 
@@
 
+The applications of the optimal transport problem to the physical transportation
+of objects is practically explicit in the problem statement. This is but one of
+many applications, though. In particular, we explore the use of optimal
+transport in the image processing of MNIST images.
+
+\begin{figure}[tb]
+  \centering
+  \includegraphics{../../code/ot-mnist-evol.pdf}
+  \caption{Optimal transport on MNIST images}
+  \label{fig:mnisttransport}
+\end{figure}
+
+Optimal transport provides a natural language with which to talk about image
+similarity. If we consider the darkness of a pixel as ``mass,''  the similarity
+of two images could be quantified as the units of mass times distance that need
+to be moved in order to transform it into the second image. This requires us to
+normalize the image such that the pixel intensities sum to 1, but this is not a
+problem with MNIST images which do not differ significantly in their average
+brightness. The transport also requires a choice of cost function between two
+pixels. We opt for an $L^2$ penalization on the distance between the pixels, but
+alternative metrics such as $L^1$ perform similarly \cref{fig:mnisttransport}
+illustrates what this transformation looks like for a ``0'' being transported
+into a ``6.''
+
+To verify that the similarity is behaving as we expect, we can compare the
+similarity of a number ``5'' against its most similar images in a small sample
+of the MNIST dataset. The most similar images are also 5's, as depicted in
+\cref{fig:mnistsim}.
+
+\begin{figure}[tb]
+  \centering
+  \includegraphics{../../code/mnist-similarity.pdf}
+  \caption{MNIST Similarity Metric}
+  \label{fig:mnistsim}
+\end{figure}
+
 \section{Conclusion} % (fold)

### 2018-04-30 reports/paper/paper.tex
./
intro

@@
 
+This project should serve as an introductory note and broad overview of both
+ theoretical and computational aspects of the
+ optimal transport literature. We hope that this note demonstrates both the
+ theoretical elegance and practical versatility of optimal transport
+ problems, and we hope to bring the reader to the forefront of the
+ literature by going through a few recent articles in detail.
+
+The organization of this report is as follows: \cref{sub:problem_intro}
+ introduces the
+ discrete optimal transport problem and discusses its interpration, dual
+ problem, and basic theoretical properties as a linear program. \cref{sec:exact} discusses methods to solve the optimal transport problem \eqref{eq:ot_problem}
+ exactly, introducing the Hungarian algorithm \cite{kuhn2010hungarian} in 
+\cref{sub:hungarian_algorithm} and presenting a proof of its correctness. The
+ Hungarian algorithm is cubic in run-time, which is optimal for exact solution
+ of \eqref{eq:ot_problem}, but is often too slow for practical
+ applications discussed in \cref{sub:applications}. As such, we discuss
+ approximate
+ optimization methods in \cref{sec:inexact_methods}, where 
+\cref{sub:entropic_regularization_and_sinkhorn_s_algorithm} focuses on the
+ Sinkhorn-Knopp algorithm used to solve a convexification of 
+\eqref{eq:ot_problem} known as \ref{eq:entropic}. We present an elegant and
+ recent upper
+ bound of Sinkhorn-Knopp's run-time by
+ \cite{altschuler2017near}, as well as numerical experiments and speculations
+ about future work. In \cref{sub:genetic_algorithms_}, we attempt to use genetic
+ algorithms to approximately solve \eqref{eq:ot_problem}, motivated by recent advances by 
+\cite{such2017deep}. In \cref{sec:applications_}, we showcase toy examples of
+ using optimal transport in computer vision with the well-known MNIST dataset.
+ \cref{sec:conclusion} concludes.
+
 \subsection{The Optimal Transport Problem and Basic Properties}
+\label{sub:problem_intro}
 
@@
 
+Recent advances in machine learning, computer vision, and statistics have
+ brought optimal transport problems to significant relevance. Specifically,
+ Wasserstein distances have received intense attention from the computer vision
+ and statistics communities. For instance, if we consider images as a discrete
+ probabilistic distribution over $\mathbb{R}^2$, the Wasserstein distance turns
+ out to mimick human perceptions of similarity: Images that are similar to
+ humans are close in Wasserstein distances (See \cite{solomon2015convolutional}
+ for an illustration of the power of Wasserstein metrics). Other distances over
+ distributions, such as the well-known Kullback-Leibler divergence, do not have
+ such properties. Such properties of the Wasserstein distance also motivates
+ advances in image generation via an incorporation with the increasingly popular
+ generative adversarial network (GAN) framework \cite{arjovsky2017wasserstein}.
+
+In terms of statistics, to fit models where likelihood evaluation is costly
+ but
+ sampling is not (e.g. with data from complicated multistage processes with
+ intractible likelihoods), we commonly sample synthetic data from the model, and
+ try to match the synthetic data with the observed data by minimizing certain
+ distances (known as Approximate Bayesian Computation in the statistics
+ literature).
+It can be shown that Wasserstein distances should be the distance of choice and
+ that Approximate Bayesian Computation via Wasserstein distances has good
+ statistical properties \cite{bernton2017inference}.
+%Thus, Wasserstein distances are extremely important in computer vision, machine learning, and statistics.
+
 % subsection applications (end)
@@
 \section{Exact Methods}
+\label{sec:exact}
 
@@
   \For{$\mathsf{generation}$ in $1:T$}
-    \For{$\mathsf{person}$ in $\mathsf{family, \gamma}$}
-      \For{$\mathsf{person}'$ in \Call{Subset}{$\mathsf{family}$}} \Comment
+    \For{$\mathsf{person}$ in $\mathsf{family}$}
+      \For{$\mathsf{person}'$ in \Call{Subset}{$\mathsf{family}, \gamma$}} \Comment
 {\textsc{Subset} chooses a subset of size $\gamma$}

### 2018-05-04 reports/paper/paper.tex
Add conclusion to paper

@@
 which makes the problem more tractable. The Hungarian algorithm solves this
-special-case.
-
-% TODO: exposition on how Hungarian works
+special-case. The Hungarian algorithm is also called a dual-ascent method
+because it starts by establishing feasible solutions to the primal and dual
+problems, and proceeds by improving these solutions until their objectives
+match. We empirically test the performance of such an algorithm, providing the
+results in \cref{fig:hungarianruntime}. As we can note when comparing to the
+general case LP, the run-times for comparably sized problems are distinctly
+smaller for the Hungarian algorithm.
 
@@
 
+To conclude, our goals in surveying the topics presented are two-fold. Firstly,
+we hope to illustrate the richness of the computational toolkit appropriate for
+solving optimal transport problems. We survey both exact methods, such as the LP
+formulation and the Hungarian method, as well as potentially faster inexact
+approximations, such as the entropic regularized problem and genetic algorithms.
+In doing so, we emphasize both theory and empirics to understand how these
+algorithms perform from different perspectives.
+
+Second, we hope to convey the flexible applicability of the optimal transport
+problem. We illustrate this with a foray into machine vision and the MNIST
+dataset as well as motivation of different intuitions for the optimal transport
+problem.
+
 \bibliographystyle{alpha}

### 2018-05-05 reports/paper/paper.tex
Modify intro to reflect Hungarian change of scope

@@
 The organization of this report is as follows: \cref{sub:problem_intro}
- introduces the
- discrete optimal transport problem and discusses its interpration, dual
- problem, and basic theoretical properties as a linear program. \cref{sec:exact} discusses methods to solve the optimal transport problem \eqref{eq:ot_problem}
- exactly, introducing the Hungarian algorithm \cite{kuhn2010hungarian} in 
-\cref{sub:hungarian_algorithm} and presenting a proof of its correctness. The
- Hungarian algorithm is cubic in run-time, which is optimal for exact solution
- of \eqref{eq:ot_problem}, but is often too slow for practical
+introduces the discrete optimal transport problem and discusses its
+interpration, dual problem, and basic theoretical properties as a linear
+program. \cref{sec:exact} discusses methods to solve the optimal transport
+problem \eqref{eq:ot_problem} exactly, introducing the Hungarian algorithm
+\cite{kuhn2010hungarian} in \cref{sub:hungarian_algorithm} and presenting an
+empirical analysis of its runtime. The Hungarian algorithm appears cubic in
+run-time, confirming theoretical asymptotic analysis, which is optimal for exact
+solution of \eqref{eq:ot_problem}, but is often too slow for practical
 applications discussed in \cref{sub:applications}. As such, we discuss
- approximate
- optimization methods in \cref{sec:inexact_methods}, where 
+approximate optimization methods in \cref{sec:inexact_methods}, where
 \cref{sub:entropic_regularization_and_sinkhorn_s_algorithm} focuses on the
@@
 \eqref{eq:ot_problem} known as \ref{eq:entropic}. We present an elegant and
- recent upper
- bound of Sinkhorn-Knopp's run-time by
- \cite{altschuler2017near}, as well as numerical experiments and speculations
- about future work. In \cref{sub:genetic_algorithms_}, we attempt to use genetic
- algorithms to approximately solve \eqref{eq:ot_problem}, motivated by recent advances by 
+recent upper bound of Sinkhorn-Knopp's run-time by \cite{altschuler2017near}, as
+well as numerical experiments and speculations about future work. In
+\cref{sub:genetic_algorithms_}, we attempt to use genetic algorithms to
+approximately solve \eqref{eq:ot_problem}, motivated by recent advances by
 \cite{such2017deep}. In \cref{sec:applications_}, we showcase toy examples of

### 2018-05-05 reports/paper/paper.tex
Fix edits

@@
  gradient methods \cite{such2017deep}. Inspired by these advances, we consider
- the following genetic algorithm
+ the following genetic algorithm.
 
@@
 The applications of the optimal transport problem to the physical transportation
-of objects is practically explicit in the problem statement. This is but one of
-many applications, though. In particular, we explore the use of optimal
-transport in the image processing of MNIST images.
+of objects is practically explicit in the problem statement. As we highlight
+previously his is but one of many applications. In particular, we explore the
+use of optimal transport in the image processing of MNIST images.
 
@@
 of two images could be quantified as the units of mass times distance that need
-to be moved in order to transform it into the second image. This requires us to
+to be moved, in order to transform it into the second image. This requires us to
 normalize the image such that the pixel intensities sum to 1, but this is not a
-problem with MNIST images which do not differ significantly in their average
+problem with MNIST images, which do not differ significantly in their average
 brightness. The transport also requires a choice of cost function between two
 pixels. We opt for an $L^2$ penalization on the distance between the pixels, but
-alternative metrics such as $L^1$ perform similarly \cref{fig:mnisttransport}
+alternative metrics such as $L^1$ perform similarly. \cref{fig:mnisttransport}
 illustrates what this transformation looks like for a ``0'' being transported
@@
 formulation and the Hungarian method, as well as potentially faster inexact
-approximations, such as the entropic regularized problem and genetic algorithms.
+approximations, such as the \eqref{eq:entropic} problem and genetic algorithms.
 In doing so, we emphasize both theory and empirics to understand how these

### 2018-05-05 reports/paper/paper.tex
fixed first half

--- 2018-05-05 reports/paper/paper.tex
@@
 This project should serve as an introductory note and broad overview of both
- theoretical and computational aspects of the
- optimal transport literature. We hope that this note demonstrates both the
+ the theoretical and the computational aspects of the
+ optimal transport literature. We hope that this note demonstrates the
  theoretical elegance and practical versatility of optimal transport
@@
 empirical analysis of its runtime. The Hungarian algorithm appears cubic in
-run-time, confirming theoretical asymptotic analysis, which is optimal for exact
+run-time, confirming theoretical asymptotic analysis; the cubic run-time 
+is also optimal for exact
 solution of \eqref{eq:ot_problem}, but is often too slow for practical
@@
 
-holds, which means that $\b P_{ij} \neq 0 \implies C_{ij} = f_i - g_j$ and
-vice versa. 
+holds, which means that $P_{ij} \neq 0 \implies C_{ij} = f_i - g_j$ and
+$C_{ij} \neq f_i - g_j \implies P_{ij} = 0$. 
 \subsection{Applications} % (fold)
@@
  Wasserstein distances have received intense attention from the computer vision
- and statistics communities. For instance, if we consider images as a discrete
- probabilistic distribution over $\mathbb{R}^2$, the Wasserstein distance turns
+ and statistics communities. For instance, if we consider an image as a discrete
+ probabilisty distribution over $\mathbb{R}^2$, where the intensity of a pixel represents density, the Wasserstein distance turns
  out to mimick human perceptions of similarity: Images that are similar to
@@
 
-The \ref{eq:ot_problem} Optimal Transport problem as formulated is an LP. This
+The \ref{eq:ot_problem} Optimal Transport problem, as formulated, is an LP. This
 means that a natural first-step to benchmarking the problem difficulty is to use
@@
 general case LP, the run-times for comparably sized problems are distinctly
-smaller for the Hungarian algorithm.
+smaller for the Hungarian algorithm. For details about the Hungarian algorithm, see \citep[Chapter 3.6]{peyre2017computational}.
 
